Appendix 1: Quantifying Drought

### 2021-01-18 analysis/paper/paper.Rmd
draft of methods section complete

*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot widely distributed throughout the Central Amazon [@kressDiversityDistributionHeliconia1990], and it is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010]. While many *Heliconia* species grow in large patches along forest edges or in disturbed areas, others---such as *H. acuminata*---are found at lower densities in shaded forest understories. Species in this group produce fewer infloresences and are pollinated by traplining hummingbirds that move from one plant to the next rather than defend a territory [@berryHeliconiaIdentificationGuide1991a]. In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@bruna2004]; reproductive plants will have 1-X <!--# from data? --> flowering shoots, each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002]. Fruits mature April-May and are consumed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011]. Dispersed seeds (1--3 seeds per fruit, $\bar{x}$=2 <!--# check data -->) germinate \~6 months after dispersal at the beginning of the rainy season, and have higher rates of germination and seedling establishment in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999]. *Heliconia acuminata* is self-incompatible and does not reproduce vegetatively.

@@

Precipitation and evapotranspiration data were obtained from a gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016]. In addition to precipitation records, this gridded dataset had estimates of potential evapotranspiration calculated using the Pennman-Monteith method. Monthly precipitation and potential evapotranspiration was used to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010]. SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale. More negative values of SPEI indicate greater drought intensity and positive values indicate wetter than average months. SPEI between 0 and -1 is considered mild drought, between -1 and -1.5 is considered moderate drought, between -1.5 and -2 is severe drought, and less than -2 is considered extreme drought [@mckee1993]. We chose a 3 month time scale as most relevant for *H. acuminata* as plants do not access sources of water buffered over longer time scales such as ground water or flood water <!--# work on this sentence -->. We calculated 3-month SPEI using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017]. We compared the SPEI values calculated using this data source to several others including near by weather stations, the global SPEI database <!--# citation --> and TRMM <!--# citation --> and the timing of drought events is consistent among these datasets although the exact magnitude of drought differs among datasets (Appendix 1)<!--# ref to appendix 1 -->.

@@

We assessed the effects of drought history on plant vital rates using distributed lag non-linear models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017]. DLNMs model potentially delayed effects of predictor variables by fitting a bi-dimensional dose-lag-response association spline, called a crossbasis function. This crossbasis function allows the effect of a predictor to vary non-linearly through lag time and across predictor intensity. Crossbasis functions were created using the *dlnm* package in R with a maximum lag of \_\_\_\_ <!--# how far back might drought impact heliconia? -->months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020]. This particular maximum lag time was determined by .... <!--# preliminary experiments? theory? limitations of data? -->. The crossbasis function was fit to the data in the context of a generalized additive model (GAM) using the *mgcv* package in R with restricted maximum likelihood [@woodGeneralizedAdditiveModels2017]. In the process of fitting the model, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity. Cubic regression splines were used for both dimensions of the crossbasis function, with \_\_\_ <!--# same as number of lag steps -1 --> knots for the lag time and 3 knots for the drought response dimension to restrict the response to drought to bimodal at most complex.

To model growth, we used shoot number as a proxy for plant size. Shoot number is easy to measure accurately and, unlike plant height, does not change if a plant is knocked over by a treefall. <!--# I could consider an integrative measure like log(shts * ht) -->. Size in year $t+1$ was modeled as a function of the natural log of size in year $t$ and drought history (i.e. the crossbasis function) with plot as a random effect. We used a scaled t error family distribution with a log link since residuals were leptokurtic when a gaussian distribution was used.

For survival, plants were assumed dead the year after they were last observed if the plant was not observed for three or more years. Survival and flowering probability were similarly modeled as a function of size and drought history (i.e. the crossbasis function) with a random effect of plot, but with a binomial family error distribution and a logit link function.

Statistical significance of crossbasis functions and covariates was determined by ....

To better understand the shape of the crossbasis function, we calculated fitted values for an average sized plant given a range of possible SPEI values from \_\_\_ to \_\_\_ at different lag times from \_\_\_ to \_\_\_ months, holding all other SPEI values at the average across the dataset (\_\_\_). These results could then be plotted as heatmaps with the *ggplot2* package ( <!--# cite ggplot2 --> ).

One limitation of the DLNM approach is that it does not currently allow for modeling of interactions with a crossbasis function. This limitation meant that we were unable to model a potential size by environment (i.e. drought history) interaction, which may occur with many plant species [@tredennickSizebyenvironmentInteractionsNeglected2018]. To check for a size by environment interaction, we followed the methods of Tredennick et al. [-@tredennickSizebyenvironmentInteractionsNeglected2018] and fit a pair of generalized linear mixed effects models (GLMMs) for each vital rate using the *lme4* package <!--# cite lme4 -->. Rather than explicitly modeling the effects of drought history, we modeled environmental stochasticity with random effects of year. For each vital rate, we included an additional random effect of year on either the intercept or both the intercept and slope of the relationship between plant size and the response. We assessed the significance of the year-by-size interaction by testing the likelihood ratio of models with and without the interaction against an expected likelihood ratio distribution under the null hypothesis of no interaction [@tredennickSizebyenvironmentInteractionsNeglected2018].

@@

@@

@@

metapopulation dynamics stuff goes here.

>>>>

*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot widely distributed throughout the Central Amazon [@kressDiversityDistributionHeliconia1990], and it is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010]. While many *Heliconia* species grow in large patches along forest edges or in disturbed areas, others---such as *H. acuminata*---are found at lower densities in shaded forest understories. Species in this group produce fewer infloresences and are pollinated by traplining hummingbirds that move from one plant to the next rather than defend a territory [@berryHeliconiaIdentificationGuide1991a]. In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@bruna2004]; reproductive plants will have 1-7 flowering shoots, each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002]. Fruits mature April-May and are consumed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011]. Dispersed seeds (1--3 seeds per fruit, $\bar{x}$=2 <!--# check data -->) germinate \~6 months after dispersal at the beginning of the rainy season, and have higher rates of germination and seedling establishment in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999]. *Heliconia acuminata* is self-incompatible and does not reproduce vegetatively.

@@

Precipitation and evapotranspiration data were obtained from a gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016]. In addition to precipitation records, this gridded dataset had estimates of potential evapotranspiration calculated using the Pennman-Monteith method. Monthly precipitation and potential evapotranspiration was used to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010]. SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale. More negative values of SPEI indicate greater drought intensity and positive values indicate wetter than average months. SPEI between 0 and -1 is considered mild drought, between -1 and -1.5 is considered moderate drought, between -1.5 and -2 is severe drought, and less than -2 is considered extreme drought [@mckeeRelationshipDroughtFrequency1993]. A 3 month time scale for SPEI was deemed as most relevant for *H. acuminata* as it relies primarily on soil moisture and not water sources buffered over longer timescales (e.g. it lacks deep taproots to access groundwater). We calculated 3-month SPEI using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017]. We compared the SPEI values calculated using this data source to several others including near by weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017] and TRMM [@trmm3b43] and the timing of drought events is consistent among these datasets although the exact magnitude of drought differs among datasets (Appendix 1).

@@

We assessed the effects of drought history on plant vital rates using distributed lag non-linear models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017]. DLNMs model potentially delayed effects of predictor variables by fitting a bi-dimensional dose-lag-response association spline, called a crossbasis function. This crossbasis function allows the effect of a predictor to vary non-linearly through lag time and across predictor intensity. Crossbasis functions were created using the *dlnm* package in R with a maximum lag of 36 months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020]. This particular maximum lag time was chosen because previous experiments showed that *H. acuminata* individuals recovered from transplant shock after 36 months, so we expected at most 36 months lag time in the response to drought [@brunaExperimentalAssessmentHeliconia2002]. The crossbasis function was fit to the data in the context of a generalized additive model (GAM) using the *mgcv* package in R with restricted maximum likelihood [@woodGeneralizedAdditiveModels2017]. In the process of fitting the model, the penalty on the crossbasis smooth (and other smoothed terms) is optimized such that more linear shapes are favored unless the data supports non-linearity.

The general form of the vital rate models was as follows:

$$
g[\textrm{E}(y_{i})] = \alpha_j + f_1(x_i) + f_2(d_{i,t}, ..., d_{i,t-24}) \\
\alpha_j \sim N(\mu_{\alpha_j}, \sigma^2_{\alpha_j}), \textrm{for plot }j
$$

Where $f_1(x_i)$ is a smooth function of plant size [ ln(height $\times$ shoot number) ], fit using a penalized cubic regression spline, $f_2()$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al. 2017 for details). Penalized cubic regression splines were used for both dimensions of the crossbasis function, with 35 knots for the lag dimension and 3 knots for the drought response dimension to restrict the fitted response to drought to bimodal at most complex. A random effect of plot ID on the intercept is represented by $a_j$. For plant growth the response was plant size (height $\times$ shoot number) in year t+1 modeled with a scaled t family error distribution and a log link function ($g(\cdot) = log(\cdot)$). A scaled t family was chosen because residuals were leptokurtic with a gaussian error structure. The data on number of infloresences was converted to binary (1 for ≥1 inflorescence, 0 for no inflorescence). For survival and flowering as a response, we used a binomial family error distribution with a logit link.

------------------------------------------------------------------------

NOTE FOR EMILIO: In Gasparrini et al. 2017, the crossbasis function is defined like so:

$$
s(x_{i,t-l_0}, ...,x_{i,t-L}) = \sum_{l = l_0}^L f \cdot w(x_{i,t-l},l) 
$$

Where "the bi-dimensional dose-lag-response function $f \cdot w (x,l)$ is composed of two marginal functions: the standard dose-response function $f(x)$, and the additional lag-response function $w(l)$ that models the lag structure...". I like this summation notation because it clarifies that the function is two-dimensional, but I'm not sure it's mathematically appropriate to drop into the model equation.\
The crossbasis function could also be represented by something a lot more abstract like $f_2(X_i, L)$ where $X_i$ is the SPEI history and $L$ is a vector (0, ..., 36) describing the lag structure of $X_i$. This is closer to how the model is written in R. Let's talk about the clearest way to write and describe these models.

------------------------------------------------------------------------

Statistical significance of crossbasis functions and covariates was determined by .... <!--# not sure yet! -->

@@

## DLNMs

@@

<!--# this all needs updating.  Probably going to use AIC instead of these p-values since I think they're doing the same thing. -->

@@

Caveats:

<!--# need to edit.  Moved from methods. -->

One limitation of the DLNM approach is that it does not currently allow for modeling of interactions with a crossbasis function. This limitation meant that we were unable to model a potential size by environment (i.e. drought history) interaction, which may occur with many plant species [@tredennickSizebyenvironmentInteractionsNeglected2018]. To check for a size by environment interaction, we followed the methods of Tredennick et al. [-@tredennickSizebyenvironmentInteractionsNeglected2018] and fit a pair of generalized linear mixed effects models (GLMMs) for each vital rate using the *lme4* package [@batesFittingLinearMixedeffects2015]. Rather than explicitly modeling the effects of drought history, we modeled environmental stochasticity with random effects of year. For each vital rate, we essentially replaced the crossbasis function in Eq 1. with a random effect of year on either the intercept or both the intercept and relationship with plant size. We assessed the significance of the year-by-size interaction by testing the likelihood ratio of models with and without the interaction against an expected likelihood ratio distribution under the null hypothesis of no interaction [@tredennickSizebyenvironmentInteractionsNeglected2018].

### 2021-01-18 analysis/supplementary-materials/appendix_SPEI.Rmd
draft of methods section complete

-   TRMM 3B43 <!--# doesn't go back far enough.  Not worth using? -->

    -   Resolution: 0.25º x 0.25º
    -   Includes: Monthly precip
    -   Doesn't include PET, so I guess I'll take that from the ground stations??
    -   Date range: 1998--2019

>>>>

-   TRMM 3B43 <!--# doesn't go back far enough.  Not worth using? -->
    -   Source DOI: 10.5067/TRMM/TMPA/MONTH/7
    -   Resolution: 0.25º x 0.25º
    -   Includes: Monthly precip
    -   Doesn't include PET, so I guess I'll take that from the ground stations??
    -   Date range: 1998--2019
